---
title: "Writing Neat Code"
author: "Randy Johnson"
date: "2/9/2017"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
library(tidyverse)

theme_set(theme_classic() +
          theme(axis.line.x = element_line(color = 'black'),
                axis.line.y = element_line(color = 'black')))
```

## Your future self

When you write code, you are collaborating with your future self.

## Reproducible Code

- Use set.seed() for random processes
- Don't ever make changes to primary data
- Maintain a clear record of what you do

## Reusable Code

- Include enough generality that you can reuse code
- Use generic, consistent variable names
- Use functions and loops
- Vectorize calculations

## Comments

- Add a short description at the top of the file
- Label sections of code
- Maintain dialog of what you are doing

## Use space

- Use tidyverse
- Use `%>%`
- Avoid packing too many things into one command
- Leave white space between lines
- Indent consistently

## Use multiple files

- Good things to put in a separate file

    - Any descrete chunk of code that gets too long
    - Functions
  
## When Sharing

- Code should be copy/pasteable

```{r eval=FALSE}
# bad code
> dat <- data_frame(x = rnorm(100),
+                   y = x + rnorm(100))
```

\textcolor{red}{\it Error: unexpected '>' in ">"}

```{r}
# good code
dat <- data_frame(x = rnorm(100),
                  y = x + rnorm(100))
```

## Loading libraries

- Either load all libraries at the top of the file
- Or include a comment on what libraries are required