---
title: "tidyverse"
author: "Randy Johnson based on a talk to the Davis R-Users' Group by Michael Levy"
date: "January 17, 2017"
output:
  pdf_document: default
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE, error = TRUE, fig.width = 4, fig.asp = 1)
```


## What is the tidyverse?

The tidyverse is a suite of R tools that follow a [tidy philosophy](https://mran.microsoft.com/web/packages/tidyverse/vignettes/manifesto.html):

### Tidy data

Put data in data frames  

- Each dataset gets a data frame
- Each variable gets a column
- Each observation gets a row

Reuse existing data structures whenever possible!

### Tidy APIs

Functions should be consistent and easily (human) readable

- Take one step at a time
- Connect simple steps with the pipe


### Okay but really, what is it? 

Suite of ~20 packages that provide consistent, user-friendly, smart-default tools to do most of what most people do in R.

- Core packages: ggplot2, dplyr, tidyr, readr, purrr, tibble
- Specialized data manipulation: hms, stringr, lubridate, forcats
- Data import: DBI, haven, httr, jsonlite, readxl, rvest, xml2
- Modeling: modelr, broom

`install.packages(tidyverse)` installs all of the above packages.

`library(tidyverse)` attaches only the core packages.


## Why tidyverse?

- Consistency  
    - e.g. All `stringr` functions take string first  
    - e.g. Many functions take data.frame first -> piping
        - Faster to write
        - Easier to read
    - Tidy data: Imposes good practices
    - Type specificity
- Implements simple solutions to common problems (e.g. `purrr::transpose`)
- Smarter defaults 
    - e.g. `utils::write.csv(row.names = FALSE)` = `readr::write_csv()` 
- Runs fast (thanks to `Rcpp`)
- Interfaces well with other tools (e.g. Spark with `dplyr` via `sparklyr`)

## `tibble`

> A modern reimagining of data frames.

```{r Attach core packages}
library(tidyverse)
```

```{r class tbl}
tdf <- tibble(x = 1:1e4, y = rnorm(1e4))   # == data_frame(x = 1:1e4, y = rnorm(1e4))
class(tdf)
```


Tibbles print politely. 

```{r print tbl}
tdf
```


- Can customize print methods with `print(tdf, n = rows, width = cols)`

- Set default with `options(tibble.print_max = rows, tibble.width = cols)`

Tibbles have some convenient and consistent defaults that are different from base R data.frames.

#### strings as factors

```{r strings as factors}
dfs <- list(df = data.frame(abc = letters[1:3], xyz = letters[24:26]),
            tbl = data_frame(abc = letters[1:3], xyz = letters[24:26])
            )
sapply(dfs, function(d) class(d$abc))
```


#### partial matching of names

```{r partial matching}
sapply(dfs, function(d) d$a)
```

#### type consistency

```{r single bracket excision}
sapply(dfs, function(d) class(d[, "abc"]))
```

Note that tidyverse import functions (e.g. `readr::read_csv`) default to tibbles and that *this can break existing code*.

#### List-columns!

```{r list columns}
tibble(ints = 1:5,
       powers = lapply(1:5, function(x) x^(1:x)))
```


## The pipe `%>%`

Sends the output of the LHS function to the first argument of the RHS function.

```{r pipe}
sum(1:8) %>%
  sqrt()
```


## `dplyr`

Common data(frame) manipulation tasks. 

Four core "verbs": filter, select, arrange, group_by + summarize, plus many more convenience functions. 


```{r load movies}
library(ggplot2movies)
str(movies)
```

```{r filter}
filter(movies, length > 360)
```

```{r select}
filter(movies, length > 360) %>%
  select(title, rating, votes)
```

```{r arrange}
filter(movies, Animation == 1, votes > 1000) %>%
  select(title, rating) %>%
  arrange(desc(rating))
```

`summarize` makes `aggregate` and `tapply` functionality easier, and the output is always a data frame.

```{r summarize}
filter(movies, mpaa != "") %>%
  group_by(year, mpaa) %>%
  summarize(avg_budget = mean(budget, na.rm = TRUE),
            avg_rating = mean(rating, na.rm = TRUE)) %>%
  arrange(desc(year), mpaa)
```


`count` for frequency tables. Note the consistent API and easy readability vs. `table`.

```{r count}
filter(movies, mpaa != "") %>%
  count(year, mpaa, Animation, sort = TRUE)
```


```{r table}
basetab <- with(movies[movies$mpaa != "", ], table(year, mpaa, Animation))
basetab[1:5, , ]
```


### joins

`dplyr` also does multi-table joins and can connect to various types of databases.

```{r full join}
t1 <- data_frame(alpha = letters[1:6],
                 num = 1:6)
t2 <- data_frame(alpha = letters[4:10],
                 num = 4:10)
full_join(t1, t2, by = "alpha", suffix = c("_t1", "_t2"))
```


## `tidyr`

Latest generation of `reshape`. `gather` to make wide table long, `spread` to make long tables wide.

```{r who}
who  # Tuberculosis data from the WHO
```

```{r gather}
who %>%
  gather(group, cases, -country, -iso2, -iso3, -year)
```


## `ggplot2`

We will cover this in more detail at a later point, but if you would like to know more, check out [one of](https://d-rug.github.io/blog/2012/ggplot-introduction) [their](https://d-rug.github.io/blog/2013/xtsmarkdown) [previous](https://d-rug.github.io/blog/2013/formatting-plots-for-pubs) [talks](https://d-rug.github.io/blog/2015/ggplot-tutorial-johnston) on ggplot or any of the excellent resources on the internet. 

Note that the pipe and consistent API make it easy to combine functions from different packages, and the whole thing is quite readable.

```{r dplyr-tidyr-ggplot}
who %>%
  select(-iso2, -iso3) %>%
  gather(group, cases, -country, -year) %>%
  count(country, year, wt = cases) %>%
  ggplot(aes(x = year, y = n, group = country)) +
  geom_line(size = .2) 
```


## `readr`

For reading flat files. Faster than base with smarter defaults.

```{r make big df}
bigdf <- data_frame(int = 1:1e6, 
                    squares = int^2, 
                    letters = sample(letters, 1e6, replace = TRUE))
```

```{r base write}
system.time(
  write.csv(bigdf, "base-write.csv")
)
```

```{r readr write}
system.time(
  write_csv(bigdf, "readr-write.csv")
)
```

```{r base read}
read.csv("base-write.csv", nrows = 3)
```

```{r readr read}
read_csv("readr-write.csv", n_max = 3)
```


## `purrr`

`purrr` is kind of like `dplyr` for lists. It helps you repeatedly apply functions. Like the rest of the tidyverse, nothing you can't do in base R, but `purrr` makes the API consistent, encourages type specificity, and provides some nice shortcuts and speed ups.

```{r intro and speedtest}
df <- data_frame(fun = rep(c(lapply, map), 2),
                 n = rep(c(1e5, 1e7), each = 2),
                 comp_time = map2(fun, n, ~system.time(.x(1:.y, sqrt))))
df$comp_time
```


### `map`

Vanilla `map` is a slightly improved version of `lapply`. Do a function on each item in a list.

```{r map}
map(1:4, log)
```

Can supply additional arguments as with `(x)apply`

```{r map arg}
map(1:4, log, base = 2)
```

Can compose anonymous functions like `(x)apply`, either the old way or with a new formula shorthand. 

```{r map formula}
map(1:4, ~ log(4, base = .x))  # == map(1:4, function(x) log(4, base = x))
```

`map` always returns a list. `map_xxx` type-specifies the output type and simplifies the list to a vector.

```{r map_type}
map_dbl(1:4, log, base = 2)
```

And throws an error if any output isn't of the expected type (which is a good thing!).

```{r map_type error}
map_int(1:4, log, base = 2)
```


`map2` is like `mapply` -- apply a function over two lists in parallel. `map_n` generalizes to any number of lists.

```{r map2}
fwd <- 1:10
bck <- 10:1
map2_dbl(fwd, bck, `^`)
```

`map_if` tests each element on a function and if true applies the second function, if false returns the original element.

```{r map_if}
data_frame(ints = 1:5, 
           lets = letters[1:5], 
           sqrts = ints^.5) %>%
  map_if(is.numeric, ~ .x^2) 
```

### Putting `map` to work

Split the movies data frame by mpaa rating, fit a linear model to each data frame, and organize the model results in a data frame.

```{r movies split models}
movies %>% 
  filter(mpaa != "") %>%
  split(.$mpaa) %>% # str()
  map(~ lm(rating ~ budget, data = .)) %>%
  map_df(tidy, .id = "mpaa-rating") %>%
  arrange(term)
```

List-columns make it easier to organize complex datasets. Can `map` over list-columns right in `data_frame`/`tibble` creation. And if you later want to calculate something else, everything is nicely organized in the data frame.

```{r list columns + map}
d <- data_frame(dist = c("normal", "poisson", "chi-square"),
                funs = list(rnorm, rpois, rchisq),
                samples = map(funs, ~.(100, 5)),
                mean = map_dbl(samples, mean),
                var = map_dbl(samples, var)
                )
d$median <- map_dbl(d$samples, median)
d
```

Let's see if we can really make this purrr... Fit a linear model of diamond price by every combination of two predictors in the dataset and see which two predict best.

```{r diamonds predictors}
train <- sample(nrow(diamonds), floor(nrow(diamonds) * .67))
setdiff(names(diamonds), "price") %>%
  combn(2, paste, collapse = " + ") %>%
  structure(., names = .) %>%
  map(~ formula(paste("price ~ ", .x))) %>%
  map(lm, data = diamonds[train, ]) %>%
  map_df(augment, newdata = diamonds[-train, ], .id = "predictors") %>%
  group_by(predictors) %>%
  summarize(rmse = sqrt(mean((price - .fitted)^2))) %>%
  arrange(rmse)
```


## `stringr`

All your string manipulation and regex functions with a consistent API. 

```{r}
library(stringr)  # not attached with tidyverse
fishes <- c("one fish", "two fish", "red fish", "blue fish")
str_detect(fishes, "two")
```

```{r}
str_replace_all(fishes, "fish", "banana")
```

```{r}
str_extract(fishes, "[a-z]\\s")
```

Let's put that string manipulation engine to work. Remember the annoying column names in the WHO data? They look like this `r stringr::str_c(colnames(tidyr::who)[5:7], collapse = ", ")`, where "new" or "new_" doesn't mean anything, the following 2-3 letters indicate the test used, the following letter indicates the gender, and the final 2-4 numbers indicates the age-class. A string-handling challenge if ever there was one. Let's separate it out and plot the cases by year, gender, age-class, and test-method.

```{r, fig.width = 8, fig.asp = .6}
who %>%
  select(-iso2, -iso3) %>%
  gather(group, cases, -country, -year) %>%
  mutate(group = str_replace(group, "new_*", ""),
         method = str_extract(group, "[a-z]+"),
         gender = str_sub(str_extract(group, "_[a-z]"), 2, 2),
         age = str_extract(group, "[0-9]+"),
         age = ifelse(str_length(age) > 2,
                      str_c(str_sub(age, 1, -3), str_sub(age, -2, -1), sep = "-"),
                      str_c(age, "+"))) %>%
  group_by(year, gender, age, method) %>%
  summarize(total_cases = sum(cases, na.rm = TRUE)) %>%
  ggplot(aes(x = year, y = total_cases, linetype = gender)) +
  geom_line() +
  facet_grid(method ~ age,
             labeller = labeller(.rows = label_both, .cols = label_both)) +
  scale_y_log10() +
  theme_light() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

