---
title: "Transformations"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(cowplot)
library(ggExtra)

# use 500 samples by default for all examples
n <- 500
```

# Transormations

In this lab we will explore different data transformations that you may find useful in helping your data meet the linearity assumption.

## Identity

`identity()` doesn't really transform the data, but we'll start here to illustrate how we will proceed with the rest of the transofmrations.

```{r identity}
# create the raw data
set.seed(293487)
dat <- data_frame(x = rnorm(n),
                  y = identity(x) + rnorm(n, sd = 0.5))

{ggplot(dat, aes(x, y)) + 
        geom_point() +
        geom_smooth() + # untransformed
        geom_smooth(method = 'lm', formula = identity(y) ~ x, color = 'orange')} %>% # transformed
    ggMarginal(type = 'histogram')
```

## Polynomial

A polynomial transformation is a sum of the variable raised to non-negative powers. The largest power that the variable is raised to is the degree of the polynomial transformation. For example, second degree polynomial transformations could be:

y + y^^2

y^^2

### Quadratic

#### `y^2` or `sqrt(x)`

```{r quadratic1, eval = FALSE}
set.seed(293478)
dat <- data_frame(x = rnorm(n, 3.5),
                  y = transformation(x + rnorm(n, sd = 0.5)))
```

#### `sqrt(y)` or `x^2`

```{r quadratic2, eval = FALSE}
set.seed(23048)
dat <- data_frame(x = rnorm(n, 4),
                  y = transformation(x) + rnorm(n, sd = 2))
```

### Cubic

#### `y^3` or `x^(1/3)`

```{r cubic1, eval = FALSE}
set.seed(234786)
dat <- data_frame(x = rnorm(n, 4),
                  y = transformation(x) + rnorm(n, sd = 0.05))
```

#### `y^(1/3)` or `x^3`

```{r cubic2, eval = FALSE}
set.seed(239487)
dat <- data_frame(x = rnorm(n),
                  y = transformation(x) + rnorm(n))
```

## Exponential

### `log(y)` or `exp(x)`

```{r exp1, eval = FALSE}
set.seed(293847)
dat <- data_frame(x = rnorm(n),
                  y = transformation(x) + rnorm(n))
```

### `exp(y)` or `log(x)`

```{r exp2, eval = FALSE}
set.seed(293847)
dat <- data_frame(x = rnorm(n, 4),
                  y = transofrmation(x) + rnorm(n, sd = 0.1))
```

## Logistic

```{r logisitc, eval = FALSE}
set.seed(239487)
dat <- data_frame(x = rnorm(n),
                  py = 1 / (1 + exp(-(x + rnorm(n, sd = 0.1)))),
                  y = rbinom(n, 1, py))

ggplot(dat, aes(x, y)) +
    geom_point() +
    geom_smooth() +
    geom_smooth(method = 'glm', formula = y ~ x, method.args = list(family = 'binomial'), color = 'orange') +
    geom_point(aes(dat$x, dat$py), color = rgb(0,0,0,.1))
```

# Practice

For each of the data sets loaded by this script, pick the best transformation for `y`.

```{r practice}
source('https://tinyurl.com/y9kau2jt')
```

* dat1: 
* dat2: 
* dat3:
* dat4: 
