---
title: "Logistic Regression"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(cowplot)
library(broom)
library(separationplot)
```

## Introduction

During this lab, we will analyze the [Medical Appointment No Shows data set from Kaggle](https://www.kaggle.com/joniarroba/noshowappointments). I've downloaded the data to this repository so we can load the data directly from this script. The data set has 15 variables and a little over 110,000 observations. It was used to find factors associated with people not showing up to meidcal appointments. See the Kaggle website if you have additional questions regarding the data, but it is fairly straight forward.

## Data summary

### Import the data

```{r import data}
# if we have already created noshow, don't do this part again - it takes forever
if(file.exists('noshow_preprocessed.RData'))
{
  load('noshow_preprocessed.RData')
}else{

  require(readr)

  if(file.exists('../Data/noshow_KaggleV2-May-2016.csv'))
  {
      noshow <- read_csv('../Data/noshow_KaggleV2-May-2016.csv')
  }else{
      noshow <- read_csv('https://github.com/BIFXapps/BIFX553/blob/master/Data/noshow_KaggleV2-May-2016.csv?raw=true')
  }

  # do we have any duplicate patient ids?
  length(unique(noshow$PatientId)) == dim(noshow)[1]
  table(table(noshow$PatientId))
  # there are ways to use all the rows of data here, but we will keep this anlysis simple and only include
  # independent observations - that is, we will only keep one appointment per individual

  # pick one appointment for each individual (this takes forever)
  set.seed(23478)
  
  noshow$keep <- NA # this is how we will track which rows to keep
  ids <- unique(noshow$PatientId) # this is all of the unique identifiers
  
  # loop over all rows in noshow (randomize order of the rows to remove possible time bias)
  for(i in sample(1:dim(noshow)[1]))
  {
    # if we haven't encountered this individual, flag the row to keep
    if(noshow$PatientId[i] %in% ids)
    {
      noshow$keep[i] <- TRUE
      ids[ids == noshow$PatientId[i]] <- ''
    }else{
      # if we have encountered this individual, don't keep this row
      noshow$keep[i] <- FALSE
    }
  }

  # drop all rows that we decided to discard above
  noshow <- filter(noshow, keep)

  # add a few additional variables that we might find useful in our anlaysis
  noshow <- mutate(noshow,
                   noshow = as.integer(`No-show` == 'Yes'),
                   scheduleLag = sapply(AppointmentDay - ScheduledDay, max, 0),
                   female = as.integer(Gender == 'F'))

  # save the dataset so we don't have to run this whole thing again!
  save(noshow, file = 'noshow_preprocessed.RData')
}
```

### Visual inspection of the data

```{r visual inspection}
ggplot(noshow, aes(Age, color = Gender)) +
  geom_histogram()

ggplot(noshow, aes(scheduleLag, color = Gender)) +
  geom_histogram()

with(noshow, table(noshow, Hipertension)) / dim(noshow)[1]
```

## Model selection

```{r model selection}
##### Models #####
model0 <- glm(noshow ~ female + scheduleLag + SMS_received + Handcap + Alcoholism +
                       Diabetes + Hipertension + Scholarship + Age + Neighbourhood,
              data = noshow, family = binomial)

model1 <- update(model0, . ~ . - Neighbourhood)


##### separation plots #####
require(separationplot)

n <- 1000
set.seed(9234)
samp <- sample(1:dim(noshow)[1], size = n)

separationplot(model0$data$noshow, predict(model0, type = 'response'), newplot = FALSE,
               lwd1 = 0.01, lwd2 = 0.01, heading = 'Full Model - all rows')
separationplot(model0$data$noshow[samp], predict(model0, type = 'response')[samp], newplot = FALSE,
               heading = 'Full Model')

##### ROC plots #####
require(pROC)

# full model - all rows
(tmp <- roc(model0$data$noshow, predict(model0, type = 'response')))
plot(tmp)
# full model - sample of rows
(tmp <- roc(model0$data$noshow[samp], predict(model0, type = 'response')[samp]))
plot(tmp)
```
